---
title: "Supplemental_Figure_1"
output: github_document
---

```{r, message=FALSE, warning=FALSE, collapse=TRUE}
library(tidyverse); library(dplyr); library(dendextend)
library(viridis); library(gplots); library(scales)

source("00_custom_functions.R")

#load files:
yeast1000 <- read_csv("../output/cleaned_data/gasch_1000_var.csv")
hbt1000 <- read_csv("../output/cleaned_data/hbt_1000_var.csv")

esr <- read_csv("../output/cleaned_data/gasch_esr.csv")
hbt <- read_csv("../output/cleaned_data/hbt_esr_percent.csv")

#whole genomes nog files:
hbt.wg <- read_csv("../data/20181113_hbt_GCF_000006805.1_ASM680v1_eggnog.csv")
sc.wg <- read_csv("../data/20181113_scer_GCF_000146045.2_R64_eggnog.csv")
colnames(hbt.wg)[1] <- "acc"
colnames(sc.wg)[1] <- "acc"

#whole genomes clusters:
clusters.yeast <- read_csv("../output/clustered_genelists/yeast2clusters.csv")
colnames(clusters.yeast)[1] <- "locus_tag"
hbt.clust <- read_csv("../output/clustered_genelists/hbt2clusters.csv")
colnames(hbt.clust)[1] <- "old_locus_tag"
```

#check that there are no duplicate gene entries and that all ESR genes are accounted for. 
```{r, message=FALSE, warning=FALSE}
length(esr$gene)
nrow(yeast1000[yeast1000$gene %in% esr$gene,]) + nrow(esr[esr$gene %ni% yeast1000$gene,])

##check that there are no duplications
#esr$gene[duplicated(esr$gene)]
#yeast1000$gene[duplicated(yeast1000$gene)]
```


##Create dendrograms and correlation matrices
analysis of top 1000 genes using spearman correlation rather than pearson
```{r, include=TRUE, results="hide"}
set.seed(123)

t.yeast1000 <- data.t(yeast1000, yeast1000$gene, scaled = "gene")
colnames(t.yeast1000) <-  colnames(t.yeast1000) %>% str_replace_all(., "\\.", "-")
yeast1000.cor <- select(t.yeast1000, -cond) %>% cor(use = "pairwise.complete.obs", method = "spearman")
yeast1000.dist <- as.dist(1-yeast1000.cor)
yeast1000.dend <- as.dendrogram(hclust(yeast1000.dist, method = "average"))

t.hbt1000 <- data.t(hbt1000, hbt1000$gene, scaled = "gene")
hbt1000.cor <- select(t.hbt1000, -cond) %>% cor(use = "pairwise.complete.obs", method = "spearman")
hbt1000.dist <- as.dist(1-hbt1000.cor)
hbt1000.dend <- as.dendrogram(hclust(hbt1000.dist, method = "average"))

##heatmap of ESR only genes, and ~16% most variable.
t.esr <- data.t(esr, esr$gene, scaled = "gene")
colnames(t.esr) <-  colnames(t.esr) %>% str_replace_all(., "\\.", "-")
esr.cor <- select(t.esr, -cond) %>% cor(use = "pairwise.complete.obs", method = "spearman")
esr.dist <- as.dist(1-esr.cor)
esr.dend <- as.dendrogram(hclust(esr.dist, method = "average"))

t.hbt <- data.t(hbt, hbt$gene, scaled = "gene")
hbt.cor <- select(t.hbt, -cond) %>% cor(use = "pairwise.complete.obs", method = "spearman")
hbt.dist <- as.dist(1-hbt.cor)
hbt.dend <- as.dendrogram(hclust(hbt.dist, method = "average"))
```

#Average expression of cluters (~16% and ESR)
```{r, collapse=TRUE, message=FALSE}
clusters.esr <- cutree(esr.dend, k=2) %>% 
  data.frame(gene = names(.), cluster = ., row.names = NULL, stringsAsFactors = F)

clusters.hbt <- cutree(hbt.dend, k=2) %>% 
  data.frame(gene = names(.), cluster = ., row.names = NULL, stringsAsFactors = F)

esr.long <- gather(t.esr, gene, expression, -cond)
hbt.long <- gather(t.hbt, gene, expression, -cond)

#set 0 to NA
esr.long[esr.long == 0] <- NA
hbt.long[hbt.long == 0] <- NA

ex.esr <- full_join(clusters.esr, esr.long, by = "gene")
ex.esr[is.na(ex.esr)] <- 0
ex.hbt <- full_join(clusters.hbt, hbt.long, by = "gene")
ex.hbt[is.na(ex.hbt)] <- 0

#get mean expression by conditon for clusters and overall
ex.esr %>%
  group_by(cluster, cond) %>%
  summarise(rep = length(expression), avg_expression = mean(expression),
            CI95 = conf_int95(expression)) -> esr.avg.ex
ex.hbt %>%
  group_by(cluster, cond) %>%
  summarise(rep = length(expression), avg_expression = mean(expression),
            CI95 = conf_int95(expression)) -> hbt.avg.ex
```

##Which cluster is induced or repressed?
```{r, include=TRUE, results="hold"}
#S. cer
clusters.esr %>%
  group_by(cluster) %>%
  tally()

clusters.yeast[clusters.yeast$locus_tag %in% filter(clusters.esr, cluster == 2)$gene,]
clusters.yeast[clusters.yeast$locus_tag %in% filter(clusters.esr, cluster == 1)$gene,]
```
cluster 2 in 1000 s. cerevisiae genes corresponds to cluster 2 in ESR only genes (same for cluster 1).

##How do esr genes cluster within entire dataset?
```{r, message=FALSE, results="hide"}
#load raw data
sc.raw <- read_csv("../output/cleaned_data/gasch_total.csv") %>% data.t(., .$gene, scaled = "gene")
colnames(sc.raw) <-  colnames(sc.raw) %>% str_replace_all(., "\\.", "-")

#make dend: (takes several minutes)
sc.cor <- select(sc.raw, -cond) %>% cor(use = "pairwise.complete.obs", method = "spearman")
sc.dist <- as.dist(1-sc.cor)
sc.dend <- as.dendrogram(hclust(sc.dist, method = "average"))
```

```{r, collapse=TRUE, message=FALSE}
sc.dend %>% #sort(type = "nodes") %>%
  raise.dendrogram (1) %>% 
  branches_attr_by_labels(labels = filter(clusters.esr, cluster == 2)$gene, attr = "col", TF_values = plasma(1, begin = 0.2)) %>%
  branches_attr_by_labels(labels = filter(clusters.esr, cluster == 1)$gene, attr = "col", TF_values = plasma(1, begin = 0.95)) %>%
  set("labels", NA) %>%
  plot(main = "ESR genes overlaid on all genes")
```

```{r, collapse=TRUE, message=FALSE}
esr1 <- yeast1000$gene %in% filter(clusters.esr, clusters.esr$cluster == 1)$gene
esr2 <- yeast1000$gene %in% filter(clusters.esr, clusters.esr$cluster == 2)$gene

rESR <- ifelse(esr1, "gold", "grey")
iESR <- ifelse(esr2, "purple", "grey")
tmp <- cbind(rESR, iESR)

yeast1000.dend %>% #sort(type = "nodes") %>%
  raise.dendrogram (1) %>% 
  color_branches(k=2, col = plasma(2, begin = 0.1, end = .7, direction = -1)) %>%
  branches_attr_by_labels(labels = filter(clusters.esr, cluster == 2)$gene, attr = "col", TF_values = plasma(1, begin = 0.2)) %>%
  branches_attr_by_labels(labels = filter(clusters.esr, cluster == 1)$gene, attr = "col", TF_values = plasma(1, begin = 0.95)) %>%
  set("labels", NA) -> dend

plot(dend, main = "ESR genes overlaid on 1,000 most variable genes") 
colored_bars(tmp, dend = dend, sort_by_labels_order = TRUE)

pdf("../output/figures/ESR_clustering.pdf")
plot(dend, main = "ESR genes overlaid on 1,000 most variable genes") 
colored_bars(tmp, dend = dend, sort_by_labels_order = TRUE)
dev.off()
```

```{r, echo=FALSE}
##Using alternative correlation methods and linking methods

#yeast1000.cor2 <- select(t.yeast1000, -cond) %>% cor(use = "pairwise.complete.obs")
#yeast1000.dist2 <- as.dist(1-yeast1000.cor2)
#yeast1000.dend2 <- as.dendrogram(hclust(yeast1000.dist2, method = "complete"))

#yeast1000.dend2 %>% #sort(type = "nodes") %>%
#  raise.dendrogram (1) %>% 
#  color_branches(k=2, col = plasma(2, begin = 0, end = .8)) %>%
#  branches_attr_by_labels(labels = filter(clusters.esr, cluster == 2)$gene, attr = "col", TF_values = plasma(1, begin = 0.25)) %>%
#  branches_attr_by_labels(labels = filter(clusters.esr, cluster == 1)$gene, attr = "col", TF_values = plasma(1, begin = 0.95)) %>%
#  set("labels", NA) -> dend

#plot(dend, main = "ESR genes overlaid on 1,000 most variable genes") 

#colored_bars(tmp, dend = dend, sort_by_labels_order = TRUE)
```

how many of iESR and rESR are in each yeast cluster?
```{r, message=FALSE, echo=FALSE}
length(intersect(filter(clusters.esr, cluster == 1)$gene, filter(clusters.yeast, cluster == 1)$locus_tag))

#length(intersect(filter(clusters.esr, cluster == 1)$gene, filter(clusters.yeast, cluster == 2)$locus_tag))
#(intersect(filter(clusters.esr, cluster == 2)$gene, filter(clusters.yeast, cluster == 1)$locus_tag))

length(intersect(filter(clusters.esr, cluster == 2)$gene, filter(clusters.yeast, cluster == 2)$locus_tag))

length(intersect(filter(clusters.esr, cluster == 1)$gene, filter(clusters.yeast, cluster == 1)$locus_tag))+length(intersect(filter(clusters.esr, cluster == 2)$gene, filter(clusters.yeast, cluster == 2)$locus_tag))
```
All ESR genes present in the top 1000 are accounted for.